Measuring the double-drift illusion with hand trajectories

### 2021-12-12 Manuscript.Rmd
suggestions from Patrick

As before, we compare the strength of the illusion (Fig. 4F), with an earlier study [@Heller2021]. A line through the origin with $K=0.60$ predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al. [-@Heller2021]. Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and organge lines are 95% confidence intervals across participants for illusion strength in each condition). While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

These first two models assumed that resets were triggered at fixed locations in space or time, but the next tro propose that the resets are random processes that occur with some distribution across space or time. We used gamma distributions to capture these processes (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants. From these likelihood values, we calculated an AIC to determine a relative probability for each model.

@@

We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018]. The results of this online tracking experiment showed a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021]. However, we were also interested in spontaneous resets and only ~4% of trials in this first experiment had deviations that could have been a reset. In contrast, in the second experiment, 63.8% of trials showed a reset. There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials. We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment if it had the same reset dynamics as the second experiment. If each pass was independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test). We are not sure why so few spontaneous resets occurred in the first experiment. We first suggest two simple explanations for the difference: 1) Online tracking required relatively fast hand movements, and therefore left little time for drawing the details of any spontaneous resets. 2) The demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020]. However, this task, unlike previous tests of the double-drift stimulus, may involve an interaction between action and perception. For example, the predicted sensory consequences of the hand movement may influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well"). Or vice versa, as in post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction"). But this is a question for future research.

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to illusion strength or stimulus onset, but may be caused by a more "random" process.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, random occurrance of distractions, a fraction of which triggers a reset, is in line with the kind of processes underlying gamma distributions [e.g. @McGill1965; @Wolfe2002].

We observe that reset times were well modeled by broad gamma distribution, suggesting that resets were best explained as occurring randomly in time with an approximate rate of 1 every 1.3 _s_. However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points. Therefore, we compare our results with those from s similar study that analyzed trajectory endpoints [pre-print: @Liu2021]. In this study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction. This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets. Anything less than that revealed the presence and size of spontaneous resets. Liu et al. tested 3 different durations (1, 2 and 3 s) and two path lengths (2 and 4 dva) for the first segment of the Gabor’s motion (see Fig. 6A). They then presented a 350 _ms_ gap during which the double-drift stimulus is not shown, and this is known to evoke a full reset to the physical location [@Lisi2015]. After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015]. By changing the location where the double-drift reappeared and asking participants if it shifted left or right, the endpoint of the first segment was determined. The difference between that point and the expected value for no spontaneous resets indicated the presence and size of any that had occurred.

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots). This can be explained by (among others) resets randomly occuring throughout time. Liu et al. [-@Liu2021] present a Poisson process model with a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds). In comparison, with the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), we found a ~0.75 / _s_ chance of a reset. Moreover, if we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without further optimization), we can also generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories. Experiment and Results based on data from Liu et al. [-@Liu2021], re-analyzed here. **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus. **B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant. These data are in line with resets occurring randomly over time. The simplest model presented by Liu et al. (2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
```

Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena. In their model, when the position signals are weak they have less weight and the motion signals make a prediction that moves the perceived position away from the true location. Their model predicts that after about 200 _ms_ the motion prediction saturates and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig. 3]. In contrast to their model, the resets in our re-tracing task show quite variable timing (Fig. 5A) and are better described by a fairly broad gamma distribution (Fig. 5B). This argues against a  mechanism that is based on comparing the current location to the physical location and saturating or triggering a reset when the discrepancy is too large. In any case, the best fitting temporal limit of the reset points in our data was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig. 5A). Moreover, 4/9 of our participants clearly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict, and only 1/9 participants showed clear "hit-the-wall" resets. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location. Our results suggest that the resets, whether saturation or returns, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration. Future models of position perception for the double-drift stimulus need to include a realistic model of the locations and type of resets.

In summary, we find that manual tracking of the double-drift stimulus during its motion did show the expected illusion at the expected strength (experiment 1). Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements could explain this too. In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a gamma distribution on reset time. This finding suggests that spontaneous resets occur independently of the stimulus and how or where it is perceived. Instead, they may occur randomly over time, perhaps due to unrelated attentional distractions.

>>>>

As before, we compare the strength of the illusion (Fig. 4F), with an earlier study [@Heller2021]. A line through the origin with $K=0.60$ predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al. [-@Heller2021]. Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and organge lines are 95% confidence intervals across participants for illusion strength in each condition). While the illusion, as captured by retracing, is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

These first two models assumed that resets were triggered at fixed locations in space or time, but the next two propose that the resets are random processes that occur with some distribution across space or time. We used gamma distributions to capture these processes (also with 2 parameters: rate; the chance of an event per second, and shape; how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants. From these likelihood values, we calculated an AIC to determine a relative probability for each model.

@@

We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018]. The results of this online tracking experiment showed a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021]. However, we were also interested in spontaneous resets and only ~4% of trials in this first experiment had deviations that could have been a reset. In contrast, in the second experiment, 63.8% of trials showed a reset. There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials. We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment if it had the same reset dynamics as the second experiment. If each pass was independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test). We are not sure why so few spontaneous resets occurred in the first experiment. We first suggest two simple explanations for the difference: 1) Online tracking required relatively fast hand movements, and therefore left little time for drawing the details of any spontaneous resets. 2) The demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020]. However, this task, unlike previous tests of the double-drift stimulus, may also involve an interaction between action and perception [e.g. @Beets2010]. For example, the predicted sensory consequences of the hand movement may influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well"). Or vice versa, as in post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction"). But this is a question for future research.

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or time from stimulus onset, but instead may be caused by a more "random" process.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of there would trigger a reset. In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.

The fit showed that reset times were well modeled by a broad gamma distribution, with resets occurring randomly in time (set by the rate and shape parameter) with an average rate of 1 reset every 1.3 _s_ over the 3 to 4 _s_ duration of the trials. However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points. To gauge the generality of our results, we compared them to those from a similar study that analyzed trajectory endpoints [pre-print: @Liu2021] to determine the presence and size of spontaneous resets. In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction. This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets. Any offset less than that revealed the presence and size of spontaneous resets. Liu et al. [-@Liu2021] tested different durations and path lengths for the first segment of the Gabor’s motion (see Fig. 6A). Then a 350 _ms_ gap in stimulus presentation is introduced, which is known to evoke a full reset to the physical location [@Lisi2015]. The difference between the perceived points where the Gabor disappeared (illusion minus spontaneous resets) and reappeared (veridical), can be used to assess spontaneous resets.

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots). This can be explained if resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint. Liu et al. [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds). In comparison, the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset. In addition, we simply used the parameters of the gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories. Experiment and Results based on data from Liu et al. [-@Liu2021], re-analyzed here. **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus. **B:** The endpoints of the first segments indicate the size of the illusion and the difference between these values and the expected value (1.4 or 2.8 _dva_) indicates the size of any intervening spontaneous resets. This shows that the size of the illusion decreases over time when path length and local illusion angle are kept constant. These data are in line with resets occurring randomly over time at ~0.7 / _s_."}
plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
```

Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena. In their model, when the position signals are weak they have less weight and the predictions based on motion signals get more weight, moving the perceived position away from the true location. Their model predicts that after about 200 _ms_ the motion prediction saturates and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig. 3]. In contrast to their model, the resets in our re-tracing task show quite variable timing (Fig. 5A) and are better described by a fairly broad gamma distribution (Fig. 5B). This argues against a  mechanism that is based on comparing the current location to the physical location and saturating or triggering a reset when the discrepancy is too large. In any case, the best fitting temporal limit of the reset points in our data was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig. 5A). Moreover, 4/9 of our participants clearly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict, and only 1/9 participants showed clear "hit-the-wall" resets. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location. Our results suggest that the resets, whether saturation or return resets, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration. Future models of position perception for the double-drift stimulus need to include a realistic model of the locations and type of resets.

In summary, we find that manual tracking of the double-drift stimulus during its motion did show the expected illusion at the expected strength (experiment 1). Few if any resets were detected during online tracking, however, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements may contribute as well. In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of experiment 2. The resets were best explained by a gamma distribution on reset time. This finding suggests that spontaneous resets occur independently of the stimulus and how or where it is perceived. Instead, they may occur randomly over time, perhaps due to unrelated attentional distractions.
